---
title: "Quickstart: commons with SAS"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown", "templates", "html_vignette", "resources", "vignette.css", package = "rmarkdown")
      - commons.css
vignette: >
  %\VignetteIndexEntry{Quickstart: commons with SAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Many teams keep their validated logic in SAS. With commons, validated SAS programs become trusted calculations, so answers built on them are marked `Verified answer`. You can also let the agent write and run its own SAS code for follow-up analysis.

This quickstart builds a small agent over clinical trial data in four steps:

1. Install SAS support and check that you can reach SAS.
2. Store the trial data in a SAS library.
3. Write a trusted SAS measure.
4. Build the agent and ask it questions.

## 1. Install and check SAS

You need an account on a SAS server. This guide uses SAS Viya 4 with home directories turned on; SAS 9.4 works too. commons reaches it through [sasquatch](https://docs.ropensci.org/sasquatch/), which runs the Python package [SASPy](https://sassoftware.github.io/saspy/) through reticulate. The commons repository is private, so pak needs a GitHub token that can read it, in `GITHUB_PAT`:

```{r}
pak::pak(c("michaelmayer2/commons/pkg-r", "sasquatch", "reticulate"))
sasquatch::install_saspy()
```

Next, tell SASPy how to connect. `configure_saspy()` opens a `sascfg_personal.py` file to edit:

```{r}
sasquatch::configure_saspy()
```

For SAS Viya 4, SASPy connects over HTTP to a compute context:

```python
SAS_config_names = ["viya"]

viya = {
    "url": "https://viya.example.com",
    "context": "SAS Studio compute context",
}
```

SASPy asks for your user name and password when it connects. A deployed app has no one to answer, so store the credentials in an `_authinfo` file and name them with `authkey`. For other ways to sign in, and for the settings SAS 9.4 needs, see the [SASPy configuration guide](https://sassoftware.github.io/saspy/configuration.html). This guide uses the configuration name `viya`.

Check that SAS runs code:

```{r}
sasquatch::sas_connect("viya")
sas <- sasquatch::sas_get_session()
cat(sas$submit("%put XCMD is %sysfunc(getoption(XCMD));")$LOG)
sasquatch::sas_disconnect()
```

The log should include `XCMD is NOXCMD`, which is SAS Viya 4's default. If it says `XCMD`, the server allows code to run host commands. Trusted measures still work, but commons won't let the agent run its own code until the server runs with `NOXCMD`. Your SAS administrator sets this. See [Security](#security).

## 2. Store the data in a SAS library

The example uses one table of trial visits: each subject's response score at baseline (week 0) and at week 12. Save it as `data/visits.csv`:

```
subject,arm,week,response
101,Placebo,0,41
101,Placebo,12,46
102,Placebo,0,38
102,Placebo,12,40
103,Placebo,0,45
103,Placebo,12,52
104,Active,0,40
104,Active,12,58
105,Active,0,36
105,Active,12,49
106,Active,0,43
106,Active,12,61
```

SAS measures read data from SAS libraries, not from R. Run this once to copy the table into a permanent library in your home directory on the SAS server, `~/trial`. The `dlcreatedir` option makes SAS create the directory if it doesn't exist:

```{r}
visits <- read.csv("data/visits.csv")

sasquatch::sas_connect("viya")
sas <- sasquatch::sas_get_session()
sas$submit("options dlcreatedir; libname trial \"~/trial\";")
sasquatch::sas_from_r(visits, "visits", libref = "trial")
sasquatch::sas_disconnect()
```

## 3. Write a trusted measure

A measure is a validated SAS program with a header that tells commons its name, what it computes, and which arguments the model may set. Save this as `measures/trial.sas`:

```sas
libname trial "~/trial" access=readonly;

/**
 * Mean response by treatment arm
 *
 * Averages the response score for each arm at one visit week.
 *
 * @measure arm_response
 * @param [week=12] `integer` Visit week: 0 for baseline, 12 for the end of treatment.
 * @return One row per arm, with the number of subjects and their mean response.
 * @output WORK.RESULT
 */
proc means data=trial.visits noprint nway;
  where week = &week;
  class arm;
  var response;
  output out=work.result(drop=_type_ _freq_) n=subjects mean=mean_response;
run;
```

How the file is read:

- The `libname` line comes before the first header, so it runs before every measure in the file.
- `~` is the home directory of the account SAS runs under, so the app must connect as the account that stored the data. To share the data with other accounts, store it in a shared directory instead, and use that path in the measure file and in `instructions`.
- `week` reaches the program as the macro variable `&week`. commons checks that the model passed an integer before the code runs.
- The dataset named by `@output` comes back to the agent as a data frame. If the log has an `ERROR` line, the measure fails, and the agent can't present the result as verified.

A file can hold several measures, each with its own header. See `?sas_measures` for the full header syntax.

## 4. Build the agent

```{r}
library(commons)

sas <- sas_session("viya")

agent <- commons(
  client = ellmer::chat("anthropic/claude-sonnet-5"),
  data_sources = list(trial = data_source(visits = read.csv("data/visits.csv"))),
  semantic_layer = semantic_layer(sas_measures("measures", session = sas)),
  sas = sas,
  instructions = paste(
    "The trial data is also in the SAS library TRIAL. In run_sas, assign it",
    "with: libname trial \"~/trial\" access=readonly;"
  )
)

commons_app(agent)
```

Each argument does one job:

- `data_sources` gives the agent the same table to describe and query with SQL.
- `sas_measures(session = sas)` runs the trusted measures on the SAS server.
- `sas = sas` gives the agent a `run_sas` tool for its own SAS code. That code runs in a second SAS session, opened with the same configuration, so it can't change the librefs, options, or macros the measures rely on. Neither session connects until it is first needed.

## Ask questions

Ask a question the measure answers:

> What was the mean response in each arm at week 12?

The agent calls `arm_response` with `week = 12`. Because a trusted measure computed the answer, it is marked `Verified answer`.

Then ask for something no measure covers:

> Did the Active arm improve more from baseline than Placebo?

The agent writes its own SAS, such as a `PROC MEANS` of each subject's change from week 0, and runs it with `run_sas`. The answer is ad hoc analysis, so it is marked `Cited` if the agent cites trusted text that supports its method, and `Untrusted` if it doesn't. Open the tool card in the chat to see the SAS code the agent ran and the output it got.

## Security

The agent's SAS code runs on the SAS server, outside the sandbox that commons uses for `run_r`. It can do anything your SAS account can, so secure the server and the account:

- **Turn off host commands.** commons only runs the agent's code if the session reports `NOXCMD`. Otherwise `run_sas` refuses and the agent tells the user that SAS analysis isn't available.
- **Keep LOCKDOWN mode on**, as SAS Viya 4 does by default. It limits SAS code to the paths your administrator allows, such as home directories.
- **Use a low-privilege account** that can read only the data the agent needs.
- **Open libraries read-only**, as the examples do with `access=readonly`.

If you only want trusted measures, leave out `sas`. Then the agent never runs its own SAS code.

## Troubleshooting

- **SASPy can't connect**: check the configuration name you passed and the settings in `sascfg_personal.py`. For SAS Viya, check `url`, `context`, and your credentials. For SAS 9.4 over IOM, check that Java is installed and on the `PATH`.
- **`run_sas is unavailable: the SAS session reports XCMD`**: ask your SAS administrator to run the server with `NOXCMD`, then run the check in step 1 again.
- **A measure ran without creating `WORK.RESULT`**: the program must create the dataset named by `@output`. Check the libref and the `output out=` statement.
