---
title: "Quickstart: commons with Altair SLC"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown", "templates", "html_vignette", "resources", "vignette.css", package = "rmarkdown")
      - commons.css
vignette: >
  %\VignetteIndexEntry{Quickstart: commons with Altair SLC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

[Altair SLC](https://altair.com/altair-slc) runs SAS-language programs without a SAS server. With commons, validated SLC programs become trusted calculations, so answers built on them are marked `Verified answer`. You can also let the agent write and run its own SAS code in SLC for follow-up analysis.

This quickstart builds a small agent over clinical trial data in four steps:

1. Install SLC support and check that SLC runs.
2. Store the trial data in an SLC library.
3. Write a trusted SLC measure.
4. Build the agent and ask it questions.

## 1. Install and check SLC

You need a licensed Altair SLC installation. commons reaches it through [slcR](https://github.com/sol-eng/slcR). The commons repository is private, so pak needs a GitHub token that can read it, in `GITHUB_PAT`:

```{r}
pak::pak(c("michaelmayer2/commons/pkg-r", "sol-eng/slcR"))
```

slcR looks for SLC in `$WPSHOME` first, then in the default installation paths such as `/opt/altair/slc/2026`. If SLC is installed somewhere else, set `WPSHOME`, for instance in `.Renviron`:

```sh
WPSHOME=/path/to/altair/slc/2026
```

Check that SLC starts and runs code:

```{r}
slc <- slcR::Slc$new()
slc$submit("%put XCMD is %sysfunc(getoption(XCMD));")
cat(slc$get_log())
slc$shutdown()
```

The log should include `XCMD is NOXCMD`. If it says `XCMD`, SLC allows code to run shell commands. Trusted measures still work, but commons won't let the agent run its own code until SLC starts with `NOXCMD`, usually set in SLC's system configuration file. See [Security](#security).

## 2. Store the data in an SLC library

The example uses one table of trial visits: each subject's response score at baseline (week 0) and at week 12. Save it as `data/visits.csv`:

```
subject,arm,week,response
101,Placebo,0,41
101,Placebo,12,46
102,Placebo,0,38
102,Placebo,12,40
103,Placebo,0,45
103,Placebo,12,52
104,Active,0,40
104,Active,12,58
105,Active,0,36
105,Active,12,49
106,Active,0,43
106,Active,12,61
```

SAS measures read data from SAS libraries, not from R. Run this once to copy the table into a permanent SLC library. This guide uses `/srv/trial`; use any directory SLC can write to:

```{r}
visits <- read.csv("data/visits.csv")

dir.create("/srv/trial", showWarnings = FALSE)
slc <- slcR::Slc$new()
slc$create_library("trial", "/srv/trial")$create_dataset_from_dataframe(
  visits,
  "visits"
)
slc$shutdown()
```

## 3. Write a trusted measure

A measure is a validated SAS program with a header that tells commons its name, what it computes, and which arguments the model may set. Save this as `measures/trial.sas`:

```sas
libname trial "/srv/trial" access=readonly;

/**
 * Mean response by treatment arm
 *
 * Averages the response score for each arm at one visit week.
 *
 * @measure arm_response
 * @param [week=12] `integer` Visit week: 0 for baseline, 12 for the end of treatment.
 * @return One row per arm, with the number of subjects and their mean response.
 * @output WORK.RESULT
 */
proc means data=trial.visits noprint nway;
  where week = &week;
  class arm;
  var response;
  output out=work.result(drop=_type_ _freq_) n=subjects mean=mean_response;
run;
```

How the file is read:

- The `libname` line comes before the first header, so it runs before every measure in the file.
- `week` reaches the program as the macro variable `&week`. commons checks that the model passed an integer before the code runs.
- The dataset named by `@output` comes back to the agent as a data frame. If the log has an `ERROR` line, the measure fails, and the agent can't present the result as verified.

A file can hold several measures, each with its own header. See `?sas_measures` for the full header syntax.

## 4. Build the agent

```{r}
library(commons)

slc <- slc_session()

agent <- commons(
  client = ellmer::chat("anthropic/claude-sonnet-5"),
  data_sources = list(trial = data_source(visits = read.csv("data/visits.csv"))),
  semantic_layer = semantic_layer(sas_measures("measures", session = slc)),
  sas = slc,
  instructions = paste(
    "The trial data is also in the SAS library TRIAL. In run_sas, assign it",
    "with: libname trial \"/srv/trial\" access=readonly;"
  )
)

commons_app(agent)
```

Each argument does one job:

- `data_sources` gives the agent the same table to describe and query with SQL.
- `sas_measures(session = slc)` runs the trusted measures on SLC.
- `sas = slc` gives the agent a `run_sas` tool for its own SAS code. That code runs in a second SLC process, so it can't change the librefs, options, or macros the measures rely on. Neither process starts until it is first needed.

## Ask questions

Ask a question the measure answers:

> What was the mean response in each arm at week 12?

The agent calls `arm_response` with `week = 12`. Because a trusted measure computed the answer, it is marked `Verified answer`.

Then ask for something no measure covers:

> Did the Active arm improve more from baseline than Placebo?

The agent writes its own SAS, such as a `PROC MEANS` of each subject's change from week 0, and runs it with `run_sas`. The answer is ad hoc analysis, so it is marked `Cited` if the agent cites trusted text that supports its method, and `Untrusted` if it doesn't. Open the tool card in the chat to see the SAS code the agent ran and the output it got.

## Security

SLC runs on the same machine as your app, under the same account. That means the agent's SAS code can read and write any file the account can:

- **Turn off shell commands.** commons only runs the agent's code if SLC reports `NOXCMD`. Otherwise `run_sas` refuses and the agent tells the user that SAS analysis isn't available.
- **Use a restricted account.** Run the app under an account that can reach only the data the agent needs.
- **Open libraries read-only**, as the examples do with `access=readonly`.

If you only want trusted measures, leave out `sas`. Then the agent never runs its own SAS code.

## Troubleshooting

- **`Could not find SLC binary`**: set `WPSHOME` to the SLC installation directory.
- **`run_sas is unavailable: the SAS session reports XCMD`**: start SLC with `NOXCMD`, then run the check in step 1 again.
- **A measure ran without creating `WORK.RESULT`**: the program must create the dataset named by `@output`. Check the libref and the `output out=` statement.
